---
title: "2. SCR_Integration_analysis"
output:
  html_document:
    df_print: paged
---
@source https://github.com/hbctraining/scRNA-seq_online/blob/master/lessons/06_SC_SCT_normalization.md
# Overview
* Check the impact of cell cycle on data.
* Evaluate effects of batch effect.
* Regress out sources of unwnated variation (nUMI, cc, mitoGenes) using Seurat SCTransform.
* Identify feature genes to integrate datasets to minimize batch effects.

## Load libraries
```{r include=TRUE, results="hide", message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(Seurat); library(tidyverse);
  library(RCurl); library(RColorBrewer);
  library(ggplot2); library(ggpubr);
  source('./helper.R')
})

```

## load data
```{r include=TRUE, results="hide", message=FALSE, warning=FALSE}
load('./data/mice_cycle.rds')
load(file="data/seurat_filtered.RData")
```

## explore cell cycle distribution - a typical source of unwanted variation
```{r include=TRUE, results="hide", message=FALSE, warning=FALSE}
seurat_phase <- NormalizeData(filtered_seurat)
seurat_phase <- CellCycleScoring(seurat_phase, g2m.features = g2m_genes, s.features = s_genes)
seurat_phase <- FindVariableFeatures(seurat_phase, selection.method = "vst", nfeatures = 3000, verbose = FALSE)
seurat_phase <- ScaleData(seurat_phase)
seurat_phase <- RunPCA(seurat_phase)
DimPlot(seurat_phase, reduction = "pca", group.by= "Phase", split.by = 'Phase')
DimPlot(seurat_phase, reduction = "pca", group.by= "sample")
DimPlot(seurat_phase, reduction = "pca", group.by= "batch")
```

## what does the umap look like w/o regress cc ?
```{r include=TRUE, results="hide", message=FALSE, warning=FALSE}
seurat_wo_integration <- NormalizeData(filtered_seurat)
seurat_wo_integration <- FindVariableFeatures(seurat_wo_integration, selection.method = "vst", nfeatures = 3000, verbose = FALSE)
seurat_wo_integration <- ScaleData(seurat_wo_integration)
seurat_wo_integration <- RunPCA(object = seurat_wo_integration, npcs = 40)

Idents(seurat_wo_integration) <- 'sample'
levels(Idents(seurat_wo_integration)) <- c('nonexpanded', 'expanded')
PCAPlot(seurat_wo_integration, dims = c(1,2))+ theme_bp
ElbowPlot(seurat_wo_integration)
```
```{r include=TRUE, results=False, message = False, echo=FALSE, warning=FALSE}
seurat_wo_integration <- RunUMAP(seurat_wo_integration, 
                             dims = 1:16,
                             reduction = "pca")
```
## w/o integration
```{r include=TRUE, results=False, message = False, echo=FALSE, warning=FALSE}
DimPlot(seurat_wo_integration,group.by = 'batch')
DimPlot(seurat_wo_integration,group.by = 'sample')

```


```{r include=TRUE, results="hide", message=FALSE, warning=FALSE}
options(future.globals.maxSize = 10000 * 1024^10)
split_seurat <- SplitObject(filtered_seurat, split.by = "orig.ident")
split_seurat
set.seed(3)
start_time <- Sys.time()
for (i in 1:length(split_seurat)) {
  split_seurat[[i]] <- NormalizeData(split_seurat[[i]], verbose =FALSE)
  split_seurat[[i]] <- CellCycleScoring(split_seurat[[i]], g2m.features=g2m_genes, s.features=s_genes)
  split_seurat[[i]] <- SCTransform(split_seurat[[i]], vars.to.regress = c("mitoRatio",'S.Score','G2M.Score'))
}

integ_features <- SelectIntegrationFeatures(object.list = split_seurat, nfeatures = 3000) 
split_seurat <- PrepSCTIntegration(object.list = split_seurat,　anchor.features = integ_features)
integ_anchors <- FindIntegrationAnchors(object.list = split_seurat,  normalization.method = "SCT", anchor.features = integ_features)
seurat_integrated <- IntegrateData(anchorset = integ_anchors,　normalization.method = "SCT")
end_time <- Sys.time()
end_time - start_time
# saveRDS(seurat_integrated, "results/seurat_integrated_mito_cc_3000_x.rds")
```
