---
title: "Slingshot"
output:
  pdf_document: default
  html_document: default
---
#Overview
* seperate lgr6 progenitor nad progenies.
* establish appropriate trajectory for both condition
* 
```{r include=TRUE, results="hide", message=FALSE, warning=FALSE}
suppressPackageStartupMessages({
  library(slingshot); library(SingleCellExperiment);library(Seurat);
  library(RColorBrewer); library(scales)
  library(viridis); library(UpSetR)
  library(pheatmap); library(msigdbr)
  library(fgsea); library(knitr)
  library(ggplot2); library(gridExtra);
  library(tradeSeq);source('helper.R')
})

```
## screen for cells that either expressing lgr6 or tdt. Total representing for lgr6 cell progenitor and progenies
```{r}
seurat_integrated <- readRDS('./results/seurat_annotated_remove_mito.rds')
DefaultAssay(seurat_integrated) <- 'RNA'
lgr6_prog <- WhichCells(seurat_integrated, slot = 'counts', expression = `tdTomato` > 0 | `Lgr6` > 0) %>% subset(seurat_integrated,cells = .)
Idents(lgr6_prog) <- 'sample'
levels(lgr6_prog) <- c('nonexpanded', 'expanded')
PCAPlot(lgr6_prog,dims = c(1,2))
```

## Integration
```{r include=TRUE, results="hide", message=FALSE, warning=FALSE}
# Load data
lgr6_prog <- readRDS('./results/lgr6_prog.rds')
saveRDS(lgr6_prog,file ='./results/lgr6_prog.rds' )
load('./data/mice_cycle.rds')
# Rename clusters
Idents(object = lgr6_prog) <- "byMarkers"
lgr6_prog$upd_anno <- Idents(RenameIdents(lgr6_prog, 
                                        "Basal"='Basal', 
                                        "Prolif_basal" ='Prolif_basal',
                                        "Spinous I"='Spinous I', 
                                        "Spinous II" ='Spinous II', 
                                        "Spinous III" ='Spinous III',
                                        "HF II" ='HF II', 
                                        "HF I" ='Infundibulum',
                                        "HF III" ='uHF',
                                        "HF IV" ='uHF',
                                        "HF V" ='HFSC',
                                        "HFSC" ='HFSC',
                                        "Immune_cell"='Immune_cell', 
                                        'SG' = "SG"))
Idents(object = lgr6_prog) <- "upd_anno"
lgr6_prog_ti <- subset(lgr6_prog, idents = c('Basal', 'uHF', 'Spinous I','Spinous II',  'Spinous III'))
```

```{r include=TRUE, results="hide", message=FALSE, warning=FALSE}
# Integration to remove technical variations
lgr6_prog_ti <- readRDS(file = './results/lgr6_prog_ti.rds')
split_seurat <- SplitObject(lgr6_prog_ti, split.by = "orig.ident")
start_time <- Sys.time()
for (i in 1:length(split_seurat)) {
  split_seurat[[i]] <- NormalizeData(split_seurat[[i]], verbose =FALSE)
  split_seurat[[i]] <- CellCycleScoring(split_seurat[[i]], g2m.features=g2m_genes, s.features=s_genes)
  split_seurat[[i]] <- SCTransform(split_seurat[[i]], vars.to.regress = c("mitoRatio",'S.Score','G2M.Score'))
}

integ_features <- SelectIntegrationFeatures(object.list = split_seurat, nfeatures = 3000) 
split_seurat <- PrepSCTIntegration(object.list = split_seurat,　anchor.features = integ_features)
integ_anchors <- FindIntegrationAnchors(object.list = split_seurat,  normalization.method = "SCT", anchor.features = integ_features)
seurat_integrated <- IntegrateData(anchorset = integ_anchors,　normalization.method = "SCT")
end_time <- Sys.time()
end_time - start_time # 11.16913 mins
seurat_integrated <- RunPCA(seurat_integrated)
seurat_integrated <- RunUMAP(seurat_integrated, dims = 1:50)
```

```{r include=TRUE, results="hide", message=FALSE, warning=FALSE}
seurat_integrated <- readRDS(file = './results/seurat_integrated_ti.rds')
DefaultAssay(seurat_integrated) <- "integrated"
DimPlot(seurat_integrated, group.by = 'byMarkers')
DimPlot(seurat_integrated, group.by = 'subcluster')
DefaultAssay(seurat_integrated) <- "RNA"
FeaturePlot(seurat_integrated, 
            features = c("Lgr6","tdTomato",'Ifitm3','Krt1','Ptgs1','Krt6b','Sparc'),
            label = FALSE, 
            sort.cell = TRUE,
            repel = TRUE
)
DefaultAssay(seurat_integrated) <- "integrated"
```

```{r include=TRUE, results="hide", message=FALSE, warning=FALSE}
# convert to singleCellExperiment
sce <- as.SingleCellExperiment(seurat_integrated, assay = "RNA")
sce$sample <- factor(sce$sample, levels = c('nonexpanded','expanded'))
sce$upd_anno <- factor(sce$upd_anno, levels = c('uHF','Basal','Spinous I','Spinous II','Spinous III'))
sce$subcluster <- factor(sce$subcluster, levels = c('Spinous', 'Basal','HF'))
```

## cell distribution
```{r include=TRUE, results="hide", message=FALSE, warning=FALSE}
# Visualize the distribution of cells in this space according to the treatment (control and TGFB) and spatial location (inner cells versus outer cells)
shuffle <- sample(ncol(sce))
layout(matrix(1:2, nrow = 1))
par(mar = c(4.5,4,1,1))

plot(reducedDims(sce)$UMAP[shuffle, ],
  asp = 1, pch = 16, xlab = "UMAP-1", ylab = "UMAP-2",
  col = alpha(c(1:2)[factor(colData(sce)$sample)][shuffle], alpha = .5))
legend("topright", pch = 16, col = 1:2, bty = "n", 
       legend = levels(factor(colData(sce)$sample)))
plot(reducedDims(sce)$UMAP[shuffle, ], asp = 1, pch = 16, xlab = "UMAP-1", ylab = "UMAP-2", 
     col = alpha(c(3:9)[factor(colData(sce)$subcluster)][shuffle], alpha = .5))
legend("topright", pch = 16, col = 3:9, bty = "n", legend = levels(factor(colData(sce)$subcluster)))
layout(1)
par(mar = c(5, 4, 4, 2) + .1)
```

## Imbalance score
```{r include=TRUE, results="hide", message=FALSE, warning=FALSE}
scores <- bioc2020trajectories::imbalance_score(
  rd = reducedDims(sce)$UMAP, 
  cl = colData(sce)$sample,
  k = 20, smooth = 40)
grad <- viridis::plasma(10, begin = 0, end = 1)
names(grad) <- levels(cut(scores$scaled_scores, breaks = 10))
plot(reducedDims(sce)$UMAP, col = grad[cut(scores$scaled_scores, breaks = 10)],
     asp = 1, pch = 16, xlab = "UMAP-1", ylab = "UMAP-2", cex = .8)
legend("topleft", legend = names(grad), col = grad, pch = 16, bty = "n", cex = 2 / 3)
```
```{r}
set.seed(42)
sce <- slingshot(sce, reducedDim = 'UMAP', clusterLabels = colData(sce)$subcluster,start.clus = 'HF', approx_points =150)
```

## order cells by progression //TODO 3 lineages
```{r plotSlingshot, echo = FALSE, fig.width =  8}
crv <- SlingshotDataSet(sce)
pseudotime <- slingPseudotime(crv, na=FALSE)
cellWeights <- slingCurveWeights(crv)

layout(matrix(c(1, 1, 2, 3), 2))
par(mar = c(4.5, 4, 1, 1))
plot(reducedDims(sce)$UMAP[shuffle, ], asp = 1, pch = 16, xlab = "UMAP-1", ylab = "UMAP-2",
  col = hcl.colors(100, alpha = .5)[cut(pseudotime, breaks = 100)][shuffle])
lines(SlingshotDataSet(sce))
# Pseudotime densities (by spatial)
ds <- list(HF = density(slingPseudotime(sce)[colData(sce)$subcluster == "HF", 1]),
           Basal = density(slingPseudotime(sce)[colData(sce)$subcluster == "Basal", 1]),
           Spinous = density(slingPseudotime(sce)[colData(sce)$subcluster == "Spinous", 1]))
xlim <- range(c(ds$HF$x, ds$Basal$x, ds$Spinous$x))
ylim <- range(c(ds$HF$y, ds$Basal$y, ds$Spinous$y))
plot(xlim, ylim, col = "white", xlab = "Pseudotime", ylab = "")
polygon(c(min(ds$Spinous$x), ds$HF$x, max(ds$HF$x)), c(0, ds$Spinous$y, 0),
        col = alpha(brewer.pal(4, "Set1")[1], alpha = .5))
polygon(c(min(ds$HF$x), ds$HF$x, max(ds$HF$x)), c(0, ds$HF$y, 0),
  col = alpha(brewer.pal(4, "Set1")[2], alpha = .5))
polygon(c(min(ds$Basal$x), ds$HF$x, max(ds$HF$x)), c(0, ds$Basal$y, 0),
        col = alpha(brewer.pal(4, "Set1")[3], alpha = .5))
legend("topleft", legend = c("Basal","HF", "Spinous") , 
       fill = alpha(brewer.pal(4, "Set1")[3:1], alpha = .5), bty = "n")

plot(reducedDims(sce)$UMAP[shuffle, ], asp = 1, pch = 16, xlab = "UMAP-1", ylab = "UMAP-2", 
     col = alpha(c(2, 3, 4)[factor(colData(sce)$subcluster)][shuffle], alpha = .5))
lines(SlingshotDataSet(sce), type = 'lineages', show.constraints = TRUE)
legend("topright", pch = 16, col = c(2,3,4), bty = "n", legend = levels(factor(colData(sce)$subcluster)))
layout(1)
par(mar = c(5, 4, 4, 2) + .1)
```
```{r plotDensities, echo=FALSE, fig.height=4}
# Pseudotime densities (by treatment)
pt <- slingPseudotime(sce)
# curve1
ne_cells <- na.omit(pt[colData(sce)$sample == "nonexpanded", 1])
e_cells <- na.omit(pt[colData(sce)$sample == "expanded", 1])
ds <- list(nonexpanded = density(ne_cells), expanded = density(e_cells))
xlim <- range(c(ds$nonexpanded$x, ds$expanded$x))
ylim <- range(c(ds$nonexpanded$y, ds$expanded$y))
plot(xlim, ylim, col = "white", xlab = "Pseudotime", ylab = "")
polygon(c(min(ds$nonexpanded$x),ds$nonexpanded$x,max(ds$nonexpanded$x)),
		c(0,ds$nonexpanded$y,0), col = rgb(0,0,0,.5))
polygon(c(min(ds$expanded$x),ds$nonexpanded$x,max(ds$nonexpanded$x)),
		c(0,ds$expanded$y,0), col = alpha(brewer.pal(4,'Set1')[1], alpha = .5))
legend("topright", legend = c("nonexpanded", "expanded"), 
       fill = alpha(c(1, brewer.pal(3, "Set1")[1]), alpha = .5), bty = "n")
```

```{r}
########################
### Kolmogorov-Smirnov Test
########################
ks.test(slingPseudotime(sce)[colData(sce)$sample == "nonexpanded", 1],
        slingPseudotime(sce)[colData(sce)$sample == "expanded", 1])

```
## Differential expression
@source https://statomics.github.io/tradeSeq/articles/fitGAM.html
```{r evaluateK, eval=FALSE}
crv <- slingCurves(sce)
pseudotime <- slingPseudotime(crv, na=FALSE)
cellWeights <- slingCurveWeights(crv)
# 5h to run, choose 5 knots
set.seed(42)
start_time <- Sys.time()
icMat <- evaluateK(counts = as.matrix(assays(sce)$counts), 
                   pseudotime = pseudotime, 
                   cellWeights = cellWeights,
                   conditions = factor(colData(sce)$sample), 
                   k=3:7, nGenes = 200, verbose = FALSE, plot = TRUE,parallel = TRUE, BPPARAM = BPPARAM)
end_time <- Sys.time()
end_time- start_time #Time difference of -1.119382 hours 1.723834 hours

# saveRDS(icMat,file = './filtered_sum/cMat_evaluatek_200_37.rds')
```
```{r}
lgr6_prog_ti <- NormalizeData(lgr6_prog_ti)
lgr6_prog_ti <- FindVariableFeatures(lgr6_prog_ti, nfeatures = 2000)
genes <- VariableFeatures(lgr6_prog_ti)
# genes <- c('Krt6b','Krt10','Krt17','Mki67','Krt14','Postn','Ifitm3','Ptgs1','Sparc','S100a8')
```

```{r}
BPPARAM <- BiocParallel::bpparam()
BPPARAM # lists current options
BPPARAM$workers <- 4 # use 4 cores
# fit gam, return fitted model
set.seed(3)
start_time <- Sys.time()
gam_sce <- fitGAM(counts = sce,
                   conditions = factor(colData(sce)$sample),
                  genes = genes,
                  nknots = 5, verbose = FALSE, parallel=TRUE, BPPARAM = BPPARAM)
end_time <- Sys.time()
start_time - end_time 
# Time difference of -1.852349 mins
# Time difference of -1.803964 hours
```

```{r}
gam_sce <- readRDS(file = './results/gam_sce.rds')
```

```{r}
mean(rowData(gam_sce)$tradeSeq$converged)
# Test gene convergence
table(rowData(gam_sce)$tradeSeq$converged)
```

## assess de along pseudotime
```{r}
rowData(gam_sce)$assocRes <- associationTest(gam_sce, lineages = TRUE, l2fc = log2(2))
assocRes <- rowData(gam_sce)$assocRes
```

```{r}
# lin1
neGenes <-  rownames(assocRes)[
  which(p.adjust(assocRes$pvalue_lineage1_conditionnonexpanded, "fdr") <= 0.05)
]
eGenes <-  rownames(assocRes)[
  which(p.adjust(assocRes$pvalue_lineage1_conditionexpanded, "fdr") <= 0.05)
]
length(neGenes)
length(eGenes)
UpSetR::upset(fromList(list(nonexpanded = neGenes, expanded = eGenes)))

```
#### Visualization of DE genes
Below we visualize and cluster the genes whose expression vary over pseudotime, using the smoothed expression patterns
As was also observed in the original manuscript, genes are mainly upregulated at the start- or endpoints of the lineage.

```{r}
## based on mean smoother lin1
yhatSmooth <- predictSmooth(gam_sce, gene = neGenes, nPoints = 50, tidy = FALSE)
heatSmooth <- pheatmap(t(scale(t(yhatSmooth[, 1:50]))),
                       cluster_cols = FALSE,
                       show_rownames = FALSE, 
                       show_colnames = FALSE)
cl <- sort(cutree(heatSmooth$tree_row, k = 6))
table(cl)

conditions <- colData(sce)$sample
pt1 <- colData(sce)$slingPseudotime_1
## based on fitted values (plotting takes a while to run)
yhatCell <- predictCells(gam_sce, gene=neGenes)
yhatCellNE <- yhatCell[,conditions == "nonexpanded"]
# order according to pseudotime
ooNE <- order(pt1[conditions == "nonexpanded"], decreasing=FALSE)
yhatCellNE <- yhatCellNE[,ooNE]
pheatmap(t(scale(t(yhatCellNE))), cluster_cols = FALSE,
          show_rownames = FALSE, show_colnames=FALSE)

```
#### Gene set enrichment analysis on genes from the NE condition
Gene set enrichment analysis on the NE genes within the mock condition confirms the biology on epithelial cell differentiation.
```{r}
# C5 category is according to gene ontology grouping: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4707969/pdf/nihms-743907.pdf
geneSets <- msigdbr(species = "Mus musculus", category = "C5", subcategory = "BP")
## filter background to only include genes that we assessed.
geneSets$gene_symbol <- toupper(geneSets$gene_symbol)
geneSets <- geneSets[geneSets$gene_symbol %in% names(sce),]
m_list <- geneSets %>% split(x = .$gene_symbol, f = .$gs_name)
stats <- assocRes$waldStat_lineage1_conditionnonexpanded
names(stats) <- rownames(assocRes)
eaRes <- fgsea(pathways = m_list, stats = stats, nperm = 5e4, minSize =1)
ooEA <- order(eaRes$pval, decreasing = FALSE)
kable(head(eaRes[ooEA, 1:3], n = 20))
```

## DE between conditions
```{r}
# outer bulge marker
plotSmoothers(gam_sce, assays(gam_sce)$counts, gene = 'Sparc',alpha = 1, border = TRUE) + ggtitle("Sparc") + theme_bw(base_line_size = 0)+
  ggsave('./ob_sparc.pdf',height = 5,width=7.5)
plotSmoothers(gam_sce, assays(gam_sce)$counts, gene = 'Cxcl14',alpha = 1, border = TRUE) + ggtitle("Cxcl14") + theme_bw(base_line_size = 0)+
  ggsave('./ob_Cxcl14.pdf',height = 5,width=7.5)
plotSmoothers(gam_sce, assays(gam_sce)$counts, gene = 'Postn',alpha = 1, border = TRUE) + ggtitle("Postn")+ theme_bw(base_line_size = 0)+
  ggsave('./ob_Postn.pdf',height = 5,width=7.5)
plotSmoothers(gam_sce, assays(gam_sce)$counts, gene = 'Krtdap',alpha = 1, border = TRUE, nPoints = 100) + ggtitle("Krtdap")+ theme_bw(base_line_size = 0)+
  ggsave('./ob_Krtdap.pdf',height = 5,width=7.5)
plotSmoothers(gam_sce, assays(gam_sce)$counts, gene = 'Cd44',alpha = 1, border = TRUE, nPoints = 100) + ggtitle("Cd44")+ theme_bw(base_line_size = 0)+theme_bw(base_line_size = 0)+
  ggsave('./ob_Cd44.pdf',height = 5,width=7.5)
#outer bulge keratinocytes characterized by expression of Krt6a/Krt75 and Cd34/Postn


#IFE/basal 
plotSmoothers(gam_sce, assays(gam_sce)$counts, gene = 'Krt14',alpha = 1, border = TRUE, nPoints = 100) + ggtitle("Krt14") + theme_bw(base_line_size = 0)+
  ggsave('./ife_krt14.pdf',height = 5,width=7.5)
plotSmoothers(gam_sce, assays(gam_sce)$counts, gene = 'Ly6e',alpha = 1, border = TRUE, nPoints = 100) + ggtitle("Ly6e")+ theme_bw(base_line_size = 0)+
  ggsave('./ife_Ly6e.pdf',height = 5,width=7.5)
plotSmoothers(gam_sce, assays(gam_sce)$counts, gene = 'Ifitm3',alpha = 1, border = TRUE, nPoints = 100) + ggtitle("Ifitm3")+ theme_bw(base_line_size = 0)+
  ggsave('./ife_Ifitm3.pdf',height = 5,width=7.5)
plotSmoothers(gam_sce, assays(gam_sce)$counts, gene = 'Mt2',alpha = 1, border = TRUE, nPoints = 100) + ggtitle("Mt2")+ theme_bw(base_line_size = 0)+
  ggsave('./ife_Mt2.pdf',height = 5,width=7.5)# predominantly IFE- derived basal cells with high expression levels of Krt14/Mt2, two stages of differentiated cells marked by Krt10/Ptgs1


plotSmoothers(gam_sce, assays(gam_sce)$counts, gene = 'Pcdh7',alpha = 1, border = TRUE, nPoints = 100) + ggtitle("Pcdh7")
plotSmoothers(gam_sce, assays(gam_sce)$counts, gene = 'Lcn2',alpha = 1, border = TRUE, nPoints = 100) + ggtitle("Lcn2") # innate immune response
plotSmoothers(gam_sce, assays(gam_sce)$counts, gene = 'S100a8',alpha = 1, border = TRUE, nPoints = 100) + ggtitle("S100a8")
plotSmoothers(gam_sce, assays(gam_sce)$counts, gene = 'S100a9',alpha = 1, border = TRUE, nPoints = 100) + ggtitle("S100a9")
plotSmoothers(gam_sce, assays(gam_sce)$counts, gene = 'S100a8',alpha = 1, border = TRUE, nPoints = 100) + ggtitle("S100a8")
plotSmoothers(gam_sce, assays(gam_sce)$counts, gene = 'Krtdap',alpha = 1, border = TRUE, nPoints = 100) + ggtitle("Krtdap")
plotSmoothers(gam_sce, assays(gam_sce)$counts, gene = 'Itgb1',alpha = 1, border = TRUE, nPoints = 100) + ggtitle("Itgb1")
plotSmoothers(gam_sce, assays(gam_sce)$counts, gene = 'Cdh1',alpha = 1, border = TRUE, nPoints = 100) + ggtitle("Cdh1")

```

```{r}
condRes <- conditionTest(gam_sce, l2fc = log2(2))
condRes$padj <- p.adjust(condRes$pvalue, "fdr")
mean(condRes$padj <= 0.05, na.rm = TRUE)
sum(condRes$padj <= 0.05, na.rm = TRUE)
conditionGenes <- rownames(condRes)[condRes$padj <= 0.05]
conditionGenes <- conditionGenes[!is.na(conditionGenes)]

```

## Visualize most and least significant gene

```{r, eval=TRUE}
# plot genes
oo <- order(condRes$waldStat, decreasing = TRUE)
# most significant gene
plotSmoothers(gam_sce, assays(gam_sce)$counts,
              gene = rownames(assays(gam_sce)$counts)[oo[1]],
              alpha = 1, border = TRUE)

```

## Heatmaps of genes DE between conditions
```{r, eval=TRUE}
# based on mean smoother
yhatSmooth <- predictSmooth(gam_sce, gene = conditionGenes, nPoints = 50, tidy = FALSE)
yhatSmoothScaled <- t(scale(t(yhatSmooth)))
heatSmooth_E <- pheatmap(yhatSmoothScaled[, 51:100],
  cluster_cols = FALSE,
  show_rownames = FALSE, show_colnames = FALSE, main = "expanded", legend = FALSE,
  silent = TRUE
)
matchingHeatmap_NE <- pheatmap(yhatSmoothScaled[heatSmooth_E$tree_row$order, 1:50],
  cluster_cols = FALSE, cluster_rows = FALSE,
  show_rownames = FALSE, show_colnames = FALSE, main = "nonexpanded",
  legend = FALSE, silent = TRUE
)
grid.arrange(heatSmooth_E[[4]], matchingHeatmap_NE[[4]], ncol = 2)

```



```{r}
statsCond <- condRes$waldStat
names(statsCond) <- rownames(condRes)
eaRes <- fgsea(pathways = m_list, stats = statsCond, nperm = 5e4, minSize = 1)
ooEA <- order(eaRes$pval, decreasing = FALSE)
kable(head(eaRes[ooEA, 1:3], n = 20))
```

